---
title: "Trias_eda"
author: "Toon Van Daele"
date: "22 augustus 2019"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    code_folding: hide
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warnings = FALSE)
library(tidyverse)
library(kableExtra)
```

# Exploratory data analysis of Trias data

This documents presents some elementary data exploration of the TRIAS dataset

The dataset originates from GBIF. See http://www.github.com/trias-project/occ-processing
for a detailed description on how the data where extracted from the GBIF database.

The data exploration uses the output from the script 'trias-test/R/mainA.R'

```{r loadrawdata}
# The RDS files are generated with the script '1_getdata.R'
#df_in <- readRDS(file = "../data/cube_belgium.RDS")
#df_bl <- readRDS(file = "../data/cube_belgium_baseline.RDS")
#df_xy <- readRDS(file = "../data/df_xy.RDS")
spec_names <- readRDS(file = "../data/spec_names.RDS")
```

## Raw data

Observations before 1950 are not relevant for the study and will be ignored.
There is no specific follow up of invasive species before 1950.

De last two years (2018 & 2019) show much less observations. This is due to the delay between
he observation in the field and the availability of the validated data in the GBIF database.
For the moment we only consider the data till 2017.

## Preprocessing

Preprocessing generates 2 data frames for the analysis:

- df_s = number of observations (obs) by cellID, year and species. The class of the species (classKey) and the number of observations of the species class (cobs).

- df_pp = Number of observations (obs) and number of distinct cells (ncells) by
year and species. The class to which the species belongs (classKey). The number of observations of the class (cobs) en number of distinct cells with observations of the class (ncobs) 


```{r load_preprocessed_data}
df_pp <- readRDS(file = "../data/df_pp.RDS")
#df_s <- readRDS(file = "../data/df_s.RDS")
```

The original dataset only contains the gridcells and years with at least one observation.
Years without observation of a species are 'gaps' in the time series.

The first year of the time series is the year with at least one observation.
Zeros are added to all consecutive years without observations till the last year (i.e. 2017).

## General overview

Description of the variables:

- obs = number of observations
- ncells = number of cells with at least 1 observaties (obs >= 1)

- cobs = number of observations of the species class
- ncobs = number of cells with observation of the species class


```{r}
nb_species <- length(unique(df_pp$taxonKey))
```

The total number of species in the dataset: `r nb_species`


Total number of observations by year (all species)

```{r observations-by-year, fig.cap="Number of species", fig.width= 5, fig.height = 3, message=FALSE, warning=FALSE}
df_pp %>%
  group_by(year) %>%
  summarise(nobs = sum(obs)) %>%
  ggplot(aes(x = year, y = nobs)) + geom_point()
```


```{r tlength}
# Length of the time series 
df_temp <- df_pp %>%
  filter(obs > 0) %>%
  group_by(taxonKey) %>%
  summarise(minyear = min(year),
            maxyear = max(year)) %>%
  group_by(minyear) %>%
  summarise(nbmin = n()) %>%
  arrange(minyear) %>%
  mutate(cumspec = cumsum(nbmin))
```

Cumulative number of species by year (first year observed)

```{r plotcumsum, fig.width= 5, fig.height = 3}
df_temp %>%
  ggplot(aes(x = minyear, y = cumspec)) + geom_line() + geom_point()
```

Number of distinct species observed since the last 10 years

```{r nbspec2017}
df_pp %>%
  filter(year > 2007, obs > 0) %>%
  group_by(year) %>%
  summarise(nbspec = n_distinct(taxonKey)) %>%
  kable()
```



```{r specyear, fig.width= 5, fig.height = 3}
df_pp %>%
  filter(obs > 0) %>%
  group_by(year) %>%
  summarise(nspec = n_distinct(taxonKey)) %>%
  arrange(year) %>%
  ggplot(aes(x = year, y = nspec)) + geom_point()
```


The number of observations and number of cells with observation in the year 2017

```{r number-of-observations-in-2017, warning=FALSE}
# df_pp %>%
#   ungroup() %>%
#   filter(year == 2017) %>%
#   arrange(obs) %>%
#   mutate(id = row_number()) %>%
#   ggplot(aes(x = id, y = obs)) + geom_point() + scale_y_log10()
```

Number of cells with observations (ncells) in 2017.

```{r number-of-cells-with-observations-in-2017, warning=FALSE}
# df_pp %>%
#   ungroup() %>%
#   filter(year == 2017) %>%
#   arrange(ncells) %>%
#   mutate(id = row_number()) %>%
#   ggplot(aes(x = id, y = ncells)) + geom_point() + scale_y_log10()
```

## Examples of some individual time series

The time series is very diverse.

Most time series are rather short with a limited number of observations per year. Often there are a lot of zeros (no observation of the species). Some examples:

```{r number-of-observations}
specs <- c("2115769", "3172100", "3686235", "4048240")
df_pp %>%
  filter(taxonKey %in% c("2115769", "3172100", "3686235", "4048240")) %>%
  left_join(spec_names, by = "taxonKey") %>%
  ggplot(aes(x = year, y = obs)) + geom_line(colour = "grey") +
    geom_point() + 
  facet_wrap(~spn, nrow = 2, scales = "free_y")
```

The number of cells (ncells) is always lower or equal to the number of observations (obs >= ncells).


