---
title: "Trias_eda"
author: "Toon Van Daele"
date: "22 augustus 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyverse)
source(file = "../R/2_preproces_data.R")
source(file = "../R/9b_plot_function.R")
```

# Exploratory data analysis of Trias data

This documents presents some basic data exploration for the TRIAS data.

The data set is a dump from GBIF -> see github -> trias-project/occ-processing

```{r load data}
# The RDS files are generated with the script '1_getdata.R'
df_in <- readRDS(file = "../data/cube_belgium.RDS")
df_bl <- readRDS(file = "../data/cube_belgium_baseline.RDS")
spec_names <- readRDS(file = "../data/spec_names.RDS")
```

## Raw data

Number of observations by year (all species)

```{r observations-by-year, fig.cap="Number of observations by year"}
df_in %>%
  group_by(year) %>%
  summarise(nobs = sum(n)) %>%
  ggplot(aes(x = year, y = nobs)) + geom_point()
```

De observaties voor 1950 laten we weg. Voor 1950 is er geen systematische opvolging
van (invasieve) soorten.

De laatste twee jaar (2018, 2019) zijn er veel minder waarnemingen. Dit is het effect van de vertraging tussen waarneming en beschikbaarheid in GBIF.
We bekijken de data voorlopig daarom slechts tot en met 2017.

## Preprocessing

Preprocessing generates 3 distinct data frames

- df_pp = Number of observations (obs) and number of distinct cells (ncells) by
year and species. The class to which the species belongs (classKey). The number of observations of the class (cobs) en number of distinct cells with observations of the class (ncobs) 

- df_s = number of observations (n) by utmID, year and species. The class of the species (classKey) and the number of observations of the class (cobs).

```{r preprocessing}
fyear <- 1950
lyear <- 2017

df_pp <- preproc(df_in, df_bl, spec_names, firstyear = fyear, lastyear = lyear)
df_s <- preproc_s(df_in, df_bl, spec_names, firstyear = fyear, lastyear = lyear)
```

The original data only contains gridcells and years with observations. Years without observation of a species are 'gaps' in the time series.

The first year of the time series is the year with at least one observation.
Zeros are added to all consecutive years without observations till the last year (i.e. 2017).

## General overview

```{r}
nb_species <- length(unique(df_pp$taxonKey))
```

The total number of species in the dataset: `r nb_species`

```{r tlength}
# Length of the time series 
df_temp <- df_pp %>%
  group_by(taxonKey) %>%
  summarise(minyear = min(year),
            maxyear = max(year)) %>%
  mutate(tlength = maxyear - minyear) %>%
  group_by(tlength) %>%
  summarise(nb = n()) %>%
  arrange(desc(tlength))

tlength0 <- df_temp %>% filter(tlength == 0) %>% .$nb
```

Number of species by lengtht of the time series (i.e. 2017 - year of first observation)

There are quite a few species with only observations for the last year (`r tlength0`) (i.e. in het jaar 2017).

```{r plottlength}
df_temp %>%
  filter(tlength > 0) %>%
  ggplot(aes(x = tlength, y = nb)) + geom_point()
```

Number of species with at least one observation

```{r specyear}
df_pp %>%
  filter(obs > 0) %>%
  group_by(year) %>%
  summarise(nspec = n_distinct(taxonKey)) %>%
  arrange(year) %>%
  ggplot(aes(x = year, y = nspec)) + geom_point()
```

Definition of the variables:

- obs = number of observations
- ncells = number of cells with at least 1 observaties (obs >= 1)

- cobs = number of observations of the species class
- ncobs = number of cells with observation of the species class

The number of observations and number of cells with observation in the year 2017

```{r number-of-observations-in-2017}
df_pp %>%
  ungroup() %>%
  filter(year == 2017) %>%
  arrange(obs) %>%
  mutate(id = row_number()) %>%
  ggplot(aes(x = id, y = obs)) + geom_point() + scale_y_log10()
```

Number of cells with observations (ncells) in 2017.

```{r number-of-cells-with-observations-in-2017}
df_pp %>%
  ungroup() %>%
  filter(year == 2017) %>%
  arrange(ncells) %>%
  mutate(id = row_number()) %>%
  ggplot(aes(x = id, y = ncells)) + geom_point() + scale_y_log10()
```

## Examples of some individual time series

The behaviour of the time series is very diverse.

Most time series are rather short with a limited number of observations per year. Many time series contain a lot of zeros (no observation of the species). Some examples:

```{r number-of-cells}
specs <- c("2115769", "3172100", "3686235", "4048240")
df_pp %>%
  filter(taxonKey %in% specs) %>%
  left_join(spec_names, by = "taxonKey") %>%
  ggplot(aes(x = year, y = ncells)) + geom_line(colour = "grey") +
    geom_point() + 
  facet_wrap(~spn, nrow = 2, scales = "free_y")
```

```{r number-of-observations}
df_pp %>%
  filter(taxonKey %in% c("2115769", "3172100", "3686235", "4048240")) %>%
  left_join(spec_names, by = "taxonKey") %>%
  ggplot(aes(x = year, y = obs)) + geom_line(colour = "grey") +
    geom_point() + 
  facet_wrap(~spn, nrow = 2, scales = "free_y")
```

The number of cells (ncells) is lower or equal to the number of observations (obs >= ncells).




